Local Type Inference for Polarised System F with Existentials

### 2023-10-12 oopsla24/tex/03-declarative-system.tex
wip

\paragraph{Quantifiers} As discussed in \cref{todo}, 
the polymorphic rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} are the only non-algorithmic ones.
For convenience of representation, we compose the left-hand side rule and
the right-hand side rule into one, and use substitution $[[σ]]$ to 
represent instantiation.

\paragraph{Substitutions} The substitution application is defined 
in a standard way, avoiding capture of bound variables,
and preserving the variables that are out of the substitution domain.
The domain and the range of the substitutions are
specified by notation $[[Γ2 ⊢ σ : Γ1]]$.
For instance, the notation $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.

\paragraph{Invariance of the Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants make the system equivalent to \systemf, 
and thus, undecidable. 
However, \ruleref{\ottdruleDOneShiftDLabel} might be
relaxed to a covariant form, as we will discuss in \cref{todo}.

@@

Another property that we extensively use is that the subtyping is reflexive and transitive. 
Moreover, any two \emph{positive} types have the least upper bound, which makes the positive subtyping a semilattice.

\begin{property}[Subtyping is well-defined]
  The subtyping is a preorder and is preserved under substitutions. 
\end{property}

>>>>

\paragraph{Quantifiers}  
Symmetric rules \ruleref{\ottdruleDOneForallLabel} and 
\ruleref{\ottdruleDOneExistsLabel} specify 
the subtyping between top-level quantified types.
Usually, the polymorphic subtyping is represented by two rules
introducing quantifiers to the left and to the right-hand side of the subtyping.
For conciseness of representation, we compose these rules into one.
First, our rule extends context $[[Γ]]$ with the quantified variables 
from the right-hand side ($[[pbs]]$ or $[[nbs]]$), 
as these variables must remain abstract.
Second, it verifies that the left-hand side quantifiers
($[[pas]]$ or $[[nas]]$) can be instantiated to continue subtyping recursively. 
The instantiation is modeled by substitution $[[σ]]$.
The notation $[[Γ2 ⊢ σ : Γ1]]$ specifies its domain and range.
For instance, $[[Γ, pbs ⊢ σ : {pas}]]$ means that 
$[[σ]]$ maps the variables from $[[pas]]$ to (positive) types
well-formed in $[[Γ, pbs]]$.
This way, application $[[ [σ]iN ]]$ instantiates (replaces) every
$[[αi⁻]]$ in $[[iN]]$ with $[[σ]]([[αi⁻]])$.

\paragraph{Invariant Shifts}
An important restriction that we put on the subtyping system is
that the subtyping on shifted types requires their equivalence,
as shown in \ruleref{\ottdruleDOneShiftDLabel} and
\ruleref{\ottdruleDOneShiftULabel}. Relaxing both of these
invariants make the system equivalent to \systemf, 
and thus, undecidable. 
However, after certain changes 
\ruleref{\ottdruleDOneShiftDLabel} can be relaxed to the covariant form,
 as we will discuss in \cref{todo}.

\paragraph{Functions}
Standardly, the subtyping of function types is covariant in the return type
and contravariant in the argument type.

\paragraph{Variables}
The subtyping of variables is defined reflexively,
which is enough to ensure the reflexivity of subtyping in general.
The algorithm will use the fact that the subtypes of a variable 
coincide with its supertypes, which however is not true for an
arbitrary type.

A property that we extensively use is that the subtyping is reflexive and transitive. 
Moreover, any two \emph{positive} types have the least upper bound, which makes the positive subtyping a semilattice.

\begin{property}[Subtyping is well-defined]
  The subtyping forms a preorder and is preserved under substitutions. 
\end{property}

\subsection{Equivalence and Normalization}

The subtyping relation forms a preorder on types,
and thus, it induces an equivalence relation \aka \emph{bicoercibility}.
The declarative specification of subtyping must be defined up to this equivalence.
Moreover, the algorithms we use must withstand changes in input types within the equivalence class.
To deal with this, we use normalization---a function that uniformly selects a representative of the equivalence class.

Using normalization gives us two benefits:
\begin{enumerate*}
  \item [(i)] we do not need to modify significantly standard operations such as unification to withstand non-trivial equivalence, and
  \item [(ii)] once the subtyping (and thus, the equivalence) changes, we only need to modify the normalization function, 
    while the rest of the algorithm remains the same.
\end{enumerate*}

\begin{figure}[h]
  \label{alg:type-nf}
  \hfill

  \begin{multicols}{2}
  \ottdefnNrmNNorm{}
  \columnbreak\\
  \ottdefnNrmPNorm{}
  \end{multicols}
  \caption{Type Normalization}
  \label{fig:type-nf}
\end{figure}

@@
